\tit

### 2021-02-19 paper.tex
Fill in missing references and explanations flagged in the text.

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated (and proved) that capture the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions, in terms of functions over lists.
\item
  A duality is revealed between regular expressions and tries in the context of language differentiation.
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than mere logical equivalence.
  As such, the differentiation lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings, including closure of languages containing the empty string.
\item
  The vocabulary and reasoning of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary.
  This generalization consists of two generally useful covariant applicative functors on functions (the usual one on functions-from-X and another on functions-to-Set).
  Properties are proved in terms of this more general vocabulary.
\item
  Properties are defined first at the level of predicates, making both easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported effortlessly to the computable parsing implementations.
\end{itemize}
The source code for this paper is freely available and contains fully verified Agda code.
All type signatures and code in the paper (including all lemmas and algorithms whether shown in the paper or elided) are generated by the Agda compiler with no postulates.

@@

\item 
\emph{Automata theory}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is the execution of a state machine.
Each state is a language, with \AB{P} being the initial state; \AF{δ} is the state transition function; and \AB{ν} is the set of accepting states \needcite{}.
Lemma \AF{ν∘foldlδ} captures the correctness of this state machine as a recognizer for the language \AB{P}.

\item
\emph{Incremental computation}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} computes \AB{P} by transforming successive incremental modifications to the empty string (each in the form of {\AB{a} \AIC ∷\_} for a list element \AB{a}) into incremental modifications to the membership question for the related language {\AF{δ} \AB{P} \AB{a}} \needcite{}.
Lemma \AF{ν∘foldlδ} again captures the correctness of this incremental implementation of \AB{P}.

\item 
\emph{Calculus}: \AB{δ} is differentiation; {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is integration; and lemma \AF{ν∘foldlδ} is the second fundamental theorem of calculus, which says that a function can be recovered by integrating its derivative \needcite{}.

@@

For an effective implementation, we will have to bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For any type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\notefoot{Explanatory footnote goes here.}
Moreover, decidability lifts naturally from types to predicates:
\agda{Decidable}
With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
Since this transformation cannot be fully automated, we will instead look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.

@@

This situation is exactly as in differential calculus, since differentiation in that setting is also defined on functions rather than on symbolic representations.
Fortunately, there are two standard solutions for \emph{computable} differentiation, going by the names of ``symbolic'' and ``automatic'' differentiation.\out{\footnote{There is also an incorrect (approximate) variation often referred to as ``numeric'' differentiation.}}
In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those representations.
The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \needcite{}.
The latter is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.

@@

Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\notefoot{Describe this derivation.}
These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type (of tries \needcite{}), while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.

@@

Looking more closely, additional algebraic structure emerges on predicates: (full) \emph{semimodules} generalize vector spaces by relaxing the associated types of ``scalars'' from fields to commutative semirings, i.e., dropping the requirements of multiplicative and additive inverses \needcite{}.
Left and right semimodules further generalize scalars to arbitrary semirings, dropping the commutativity requirement for scalar multiplication.
Again, nothing is needed from the codomain type \AB{B} beyond the properties of a semiring, so again predicates get these algebraic structures for free (already paid for by types).
For every {\AB{P} \AS : \AF{Pred} \AB{A}}, \AB{P} is a vector in the semimodule {\AF{Pred} \AB{A}}, including the special case of ``languages'', for which \AB{A} is a list type.
The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling conjoining membership proofs.

There is still more algebraic structure to be found and exploited.
When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \needcite{}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Specifying Languages}.
In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \needcite{}.
The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Specifying Languages}.
The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations, applied to any domain monoid.
The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
The proofs of these properties, while reasonably straightforward, are beyond the scope of this paper but are available in the paper's Agda source code.\out{\notefoot{There will be short and long forms of this paper. Perhaps the long form will contain narrated parts of the monoid semiring proof.}}

@@

Most work on formal languages in functional programming has been in parsing, particularly via parsing combinators \needcite{}.
Such work typically concentrates on usability, composability, and sometimes efficiency, rather than on simple, precise specification and verifiable correctness.

>>>>

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated (and proved) about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions.
\item
  A duality is revealed between regular expressions and tries in the context of language differentiation.
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than mere logical equivalence.
  As such, the differentiation lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings, including closure of languages containing the empty string.
\item
  The vocabulary and reasoning of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary.
  This generalization consists of two generally useful covariant applicative functors on functions (the usual one on functions-from-X and another on functions-to-Set).
  Properties are proved in terms of this more general vocabulary.
\item
  Properties are defined first at the level of predicates, making both easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported effortlessly to the computable parsing implementations.
\end{itemize}
The source code for this paper is freely available and contains fully verified Agda code.
All type signatures and code in the paper (including all lemmas and algorithms whether shown in the paper or elided) are generated by the Agda compiler with no postulates.

@@

\item 
\emph{Automata theory}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is the execution of a state machine.
Each state is a language, with \AB{P} being the initial state; \AF{δ} is the state transition function; and \AB{ν} is the set of accepting states\out{ \needcite{}}.
Lemma \AF{ν∘foldlδ} captures the correctness of this state machine as a recognizer for the language \AB{P}.

\item
\emph{Incremental computation}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} computes \AB{P} by transforming successive incremental modifications to the empty string (each in the form of {\AB{a} \AIC ∷\_} for a list element \AB{a}) into incremental modifications to the membership question for the related language {\AF{δ} \AB{P} \AB{a}}\out{ \needcite{}}.
Lemma \AF{ν∘foldlδ} again captures the correctness of this incremental implementation of \AB{P}.

\item 
\emph{Calculus}: \AB{δ} is differentiation; {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is integration; and lemma \AF{ν∘foldlδ} is the second fundamental theorem of calculus, which says that a function can be recovered by integrating its derivative\out{ \needcite{}}.

@@

For an effective implementation, we will have to bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For any type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
A few explanatory remarks:
\begin{itemize}

\item {\AS λ \AS{()}} is the unique function from \AD ⊥ to any type (here also {\AD ⊥}).

\item \AF{inj₁} and \AF{inj₂} are left and right injections into a sum type.

\item \AF{proj₁} and \AF{proj₂} are left and right projections out of a product type.

\item Given two functions {\AB f \AS : \AB A \AS → \AB C} and {\AB g \AS : \AB B \AS → C}, the function {\AF [ \AB f \AF , \AB g \AF ] \AS : \AB A \AF ⊎ \AB B \AS → \AB C} maps {\AIC{inj₁} \AB a} to {\AB f \AB a} and {\AIC{inj₂} \AB b} to {\AB g \AB b}.
In this use, {\AB C} is \AD ⊥, since we are proving a negation.

\item The \AF{\_✶‽} definition reflects the fact that the empty list exists for every element type (even \AF ⊥).

}
Moreover, decidability lifts naturally from types to predicates:
\agda{Decidable}
With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
Since this transformation cannot be fully automated, we will instead look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.

@@

This situation is exactly as in differential calculus, since differentiation in that setting is also defined on functions rather than on symbolic representations.
Fortunately, there are two standard solutions for \emph{computable} differentiation, going by the names of ``symbolic'' and ``automatic'' differentiation.\out{\footnote{There is also an incorrect (approximate) variation often referred to as ``numeric'' differentiation.}}
In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those representations.
The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
The latter is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.

@@

Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{The classic trie (``prefix tree'') data structure was introduced by Thue in 1912 to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary (regular) algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.

@@

Looking more closely, additional algebraic structure emerges on predicates: (full) \emph{semimodules} generalize vector spaces by relaxing the associated types of ``scalars'' from fields to commutative semirings, i.e., dropping the requirements of multiplicative and additive inverses.
Left and right semimodules further generalize scalars to arbitrary semirings, dropping the commutativity requirement for scalar multiplication.
Again, nothing is needed from the codomain type \AB{B} beyond the properties of a semiring, so again predicates get these algebraic structures for free (already paid for by types).
For every {\AB{P} \AS : \AF{Pred} \AB{A}}, \AB{P} is a vector in the semimodule {\AF{Pred} \AB{A}}, including the special case of ``languages'', for which \AB{A} is a list type.
The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling conjoining membership proofs.

There is still more algebraic structure to be found and exploited.
When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \citep{Golan2005RecentSemi}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Specifying Languages}.
In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \cite{Golan2005RecentSemi, Elliott2019-convolution}.
The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Specifying Languages}.
The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations, applied to any domain monoid.
The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
The proofs of these properties, while reasonably straightforward, are beyond the scope of this paper but are available in the paper's Agda source code.\out{\notefoot{There will be short and long forms of this paper. Perhaps the long form will contain narrated parts of the monoid semiring proof.}}

@@

Most work on formal languages in functional programming has been in parsing, particularly via parsing combinators \citep{Hutton1996monadic, Leijen2001parsec, Swierstra2008combinator}.
Such work typically concentrates on usability, composability, and sometimes efficiency, rather than on simple, precise specification and verifiable correctness.
